---
title: "Data Analyst Job Postings: A Brief Report"
author: "Tamayi Mlanda"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)

library(ggplot2)
library(dplyr)
library(readr)
library(knitr)

postings <- read_csv("data/data_analyst_postings_sample.csv", show_col_types = FALSE)

postings <- postings |>
  mutate(
    `Minimum Years Experience` = suppressWarnings(as.numeric(`Minimum Years Experience`)),
    `Median Salary Estimate` = suppressWarnings(as.numeric(`Median Salary Estimate`)),
    `Data Languages` = factor(`Data Languages`, levels = c("R", "Python", "Both", "Neither"))
  )

n_total <- nrow(postings)
n_salary <- sum(!is.na(postings$`Median Salary Estimate`))
n_exp <- sum(!is.na(postings$`Minimum Years Experience`))
n_lang <- sum(!is.na(postings$`Data Languages`))
```

## Introduction

This report explores a sample of Data Analyst job postings, focusing on how
**minimum years of experience** and **required data languages** relate to the
**median salary estimate** advertised in each posting.

The dataset contains `r format(n_total, big.mark = ",")` postings in total, of
which `r format(n_salary, big.mark = ",")` have a usable salary estimate,
`r format(n_exp, big.mark = ",")` have a parseable minimum-experience value, and
`r format(n_lang, big.mark = ",")` are tagged with a data-language category.
Because experience and language requirements are extracted from free-text job
descriptions, missing or noisy values are expected.

## Experience vs. Salary

```{r scatter}
plot_df <- postings |>
  filter(!is.na(`Minimum Years Experience`), !is.na(`Median Salary Estimate`))

ggplot(plot_df, aes(x = `Minimum Years Experience`, y = `Median Salary Estimate`)) +
  geom_point(alpha = 0.4, colour = "#2C7FB8") +
  geom_smooth(method = "lm", se = TRUE, colour = "#D95F0E", fill = "#FED98E") +
  scale_y_continuous(labels = scales::label_dollar()) +
  labs(
    x = "Minimum Years of Experience",
    y = "Median Salary Estimate (USD)",
    title = "Required Experience vs. Median Salary Estimate"
  ) +
  theme_minimal(base_size = 12)
```

**Interpretation.** The scatter plot suggests a modest positive association
between the minimum years of experience a posting requires and its median
salary estimate: postings asking for more experience tend, on average, to
advertise higher pay. The relationship is far from tight — there is wide
salary variation at every experience level, reflecting differences in
industry, location, company size, and seniority that this single variable
does not capture.

## Salary by Data Language

```{r boxplot}
box_df <- postings |>
  filter(!is.na(`Data Languages`), !is.na(`Median Salary Estimate`))

ggplot(box_df, aes(x = `Data Languages`, y = `Median Salary Estimate`, fill = `Data Languages`)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.3) +
  scale_y_continuous(labels = scales::label_dollar()) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(
    x = "Data Languages Mentioned",
    y = "Median Salary Estimate (USD)",
    title = "Median Salary Estimate by Data Language Requirement"
  ) +
  theme_minimal(base_size = 12)
```

```{r lang-table}
box_df |>
  group_by(`Data Languages`) |>
  summarise(
    Postings = n(),
    `Median Salary` = scales::label_dollar()(median(`Median Salary Estimate`, na.rm = TRUE)),
    `Mean Salary`   = scales::label_dollar()(mean(`Median Salary Estimate`, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  kable(caption = "Salary summary by data-language requirement")
```

**Interpretation.** Postings that mention **Python**, or **both Python and R**,
tend to advertise somewhat higher median salaries than those mentioning only
**R** or **neither** language. The boxes overlap considerably, however, so the
language category alone is not a strong predictor of pay — it is more likely a
proxy for the kind of role (e.g. management vs. analytics engineering vs. reporting) than a
direct salary driver.

## Limitations

- **Sample size and balance.** Although the file contains many rows, the
  number of postings with *all three* fields populated is much smaller, and
  the language categories are unevenly represented — a few categories carry
  most of the data, which makes group comparisons fragile.
- **Noisy extraction.** Both "Minimum Years Experience" and "Data Languages"
  are derived from free-text descriptions. Postings where extraction failed
  are dropped, which may bias the remaining sample toward more structured or
  explicitly-written ads.
- **Salary is an estimate.** "Median Salary Estimate" is itself a derived
  figure, not a verified offer, so absolute dollar values should be read as
  indicative rather than precise.
- **Omitted variables.** Location, industry, and company size all influence
  pay and are not controlled for in these two plots.

## Personal Reflection

- **LLMs Make Data Analysis Accessible:** Given the LLM, we were able to 
  produce to a polished report with minimal coding and data manipulation. 
  This opens up opportunities for people without advanced programming skills
  to be able to quickly produce summary reports.
- **LLMs Speed Up Data Analysis Tasks:** By relying on the LLM to analyse
  the text data of the Job Descriptions (natural language), we were able to
  quickly extract data and generate insights from the job descriptions. Previously,
  such a task would have had to be done manually, and that would have been time
  consuming because we would have to read through each job description.
- **LLM Risks:** 
  - **Implicit Trust in the LLM, it might not always get it right:** The LLM is 
    not perfect, and it can make mistakes in data extraction and
    analysis. It is important to verify the results produced by the LLM and to
    be aware of its limitations. For example, the LLM may have difficulty
    accurately extracting data from unstructured text, and it may not be able
    to capture all the nuances of the job descriptions. In this case, we did not 
    verify whether it had extracted all the data correctly, we trusted it.
  - **The LLM is a black box, and inexperienced data analysts might not be able** 
    **to interrogate the outputs of an LLM:** Because the LLM is a black box, it can 
    be difficult to understand how it arrived at its conclusions. This can make it 
    challenging to identify and correct any errors in the analysis. It also makes 
    it difficult to understand the reasoning behind the insights generated by the 
    LLM, which can be a problem when trying to communicate those insights to others.
- **My Prompts:**  

  Exract Minimum Years of Experience:
  ```
  =MYGPT(CONCATENATE(""Extract the minimum years of experience required from this job description. Return ONLY a number. If a range is given, return the lower number. If no years of experience are explicitly mentioned, return 'NA'. Here's the job description: "", C2),GEN_AI_MODEL_NAME,OPENROUTER_API_KEY)
  ```
    
  Extract Programming Language Requirement:
  ```
  =MYGPT(CONCATENATE("Analyze the Job Description and return ONLY one of these four specific strings based on the data programming language requirements in the job description: R, Python, Both, or Neither. Here's the job description: ", C2), GEN_AI_MODEL_NAME, OPENROUTER_API_KEY)
  ```